Component response rate variation underlies the stability of complex systems

### 2020-02-23 notebook/ms.Rmd
Add new sentence to the introduction listing robustness of network types, and add associated bibtex entries.

Author Summary
--------------------------------------------------------------------------------

Any set of discrete components that potentially interact can define a complex system. Typically, properties of individual components will cause them to respond to system perturbation at different rates, but the consequences of such variation for system stability are unexplored. This work makes three novel theoretical contributions: First, I show that large complex systems with varying component response rates are more likely to be stable than when response rates are uniform. Second, I show that a targeted manipulation of component response rates can increase the probability of system stability by several orders of magnitude. Lastly, I show that varying component response rate does not affect system feasibility. These results are general to any physical, biological, social, or mixed system.

@@

May's [@May1972; @Allesina2012] stability criterion $\sigma\sqrt{SC} < 1$ assumes that the expected response rates ($\gamma$) of individual components to perturbations of the system are identical, but this is highly unlikely in any complex system. In ecological communities, for example, the rate at which population density changes following perturbation will depend on the generation time of organisms, which might vary by orders of magnitude among species. Species with short generation times will respond quickly (high $\gamma$) to perturbations relative to species with long generation times (low $\gamma$). Similarly, the speed at which individual banks respond to perturbations in financial networks, or individuals or institutions respond to perturbations in complex social networks, is likely to vary. The effect of such variance on stability has not been investigated in complex systems theory. Intuitively, variation in $\gamma$ might be expected to decrease system stability by introducing a new source of variation into the system and thereby increasing $\sigma$. Here I show why, despite higher $\sigma$, realistic complex systems (such that $S$ is high but finite) are actually more likely to be stable if their individual component response rates vary. <!--- Something about including realistic structured networks (scale free, small world, and cascade food webs) --->

>>>>

@@

May's [@May1972; @Allesina2012] stability criterion $\sigma\sqrt{SC} < 1$ assumes that the expected response rates ($\gamma$) of individual components to perturbations of the system are identical, but this is highly unlikely in any complex system. In ecological communities, for example, the rate at which population density changes following perturbation will depend on the generation time of organisms, which might vary by orders of magnitude among species. Species with short generation times will respond quickly (high $\gamma$) to perturbations relative to species with long generation times (low $\gamma$). Similarly, the speed at which individual banks respond to perturbations in financial networks, or individuals or institutions respond to perturbations in complex social networks, is likely to vary. The effect of such variance on stability has not been investigated in complex systems theory. Intuitively, variation in $\gamma$ might be expected to decrease system stability by introducing a new source of variation into the system and thereby increasing $\sigma$. Here I show that, despite higher $\sigma$, realistic complex systems (such that $S$ is high but finite) are actually more likely to be stable if their individual component response rates vary. My results are robust across commonly observed network structures, including random [@May1972], small-world [@Watts1998], scale-free [@Albert2002], cascade food web [@Williams2000] networks.
